Psychophysics of musical elements in the discrete-time representation of sound

### 2013-10-06 article/article.tex
feito ateh modulacao

The simultaneous occurrence of three or more notes is observed by means of chords. Their
base of tonal music are the triads. Triads are built by two successive thirds
within 3 notes: root, third and fifth. If the lower note of a chord is different from the root, this chord is referred
as inverted. A close position is where any note of chord fits between two consecutive
notes~\cite{Lacerda}. All the non-inverted triads in the close position form and with root in
$0$ are described as following:

@@

To consider another third superimposed to the fifth, is sufficient to add $10$
as the highest note in order to form a tetrad with minor seventh, or add $11$ in order to form a tetrad with major
seventh. The inversions and open positions can be obtained with a selective
addition of $12$ to the components.

@@

In the following we present general recommendations:

\begin{itemize}
        \item A fifth is a root (fundamental) confirmed by the interval.
        \item The major or minor third points to the major or minor quality
        of the chord.
        \item Every tritone, specially if built between a major third and a
        minor seventh, tends to resolve in a third or sixth.
        \item Note duplication is avoided. If duplication is needed, the
        preferred order is: the root, fifth, third and seventh.
        \item It is possible to build chords with notes different from triads,
        specially if they respect to a recurrent logic or musical chain that
        justifies these different notes.
        \item Chords built by successive intervals different of thirds -- like
        fourths and seconds -- are recurrent in compositions with advanced
        tonalism or experimental character.
        \item The repetition of chords successions (or their characteristics)
        fixes a trajectory by means of the recurrence and make possible the
        introduction of exotic formations without incoherence.
\end{itemize}

@@

Omission of the basic chaining is the key to obtain modal and atonal harmonies. In
case of absence of these minimal tonal structures, we say that the
harmony is modal if the notes match with some diatonic scale (see
equations~\ref{eq:escalas}) or if the notes are presented in a small number.
If basic tonal chaining are absent, the notes do not match any diatonic scale
and are diverse and dissonant (by relation with each other) enough to avoid
reduction by polarization, the harmony is atonal. In this classification, the
modal harmony are not tonal neither atonal. The modal harmony is reduced to the 
incidence of notes within the diatonic scale (or simplifications) and to the
absence of tonal structures. It is possible to note, following this concept, that
atonal harmony is hard to be realized~\cite{harmEXT}.

@@

In fact, the techniques around atonal music comprehend structures for avoiding
the hearing relation with modes and tonality. This difficulty imposes the use of dodecafonism. 
The purpose of dedocafonism is to use a set of notes (ideally 12 notes), and to perform each note, one by one, in the same
order (não sei se ficou mt claro). In this context, the tonic becomes difficult to settle. Nevertheless, the
Western hearing searches for tonal traces in musics and easily finds them by
unexpected paths and even tortuous.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The use of dissonant intervals (specially tritones without resolution), together
with ninths, seconds and sevenths, reinforces the absence of tonality. Considering
this context, it is possible to account for the following considerations while creating a music
piece:

\begin{itemize}
     \item Repeating notes. Considering the immediate repetition as an extension
     of the previous incidence, the use of same notes in sequence does not add relevant information.
     \item It is interesting to play adjacent notes at the same time, making harmonic intervals
     and chords.
     \item Present durations with liberty, respecting the notes order.
     \item In order to gain variation, one can use extension, transposition, translation, inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
     and~\ref{subsec:usosmusicais3} for more details.
     \item Account for the presence of note structures, it is possible to use variations in orchestration, articulation, spatialization, among others.
\end{itemize}

The atonal harmony can be observed, paradigmatically, within these presented
conditions. Most of them were written by great dodecafonic composers like Alban Berg and even Schoenberg. Many pieces create purposeful links between tonal and atonal techniques.

@@

In the XX century, rhythmic music with emphasis on sonorities/timbres, extended
the concepts about tonality and harmony. Nevertheless, the
tonal harmony has strong influence in the artistic movements and commercial
venues. In addition, the dodecafonism is considered having tonal nature because it
was created to deny the tonal characteristics of polarization.

In tonal or modal music, the base of the harmonic filed are chords -- like the ones listed in
equations~\ref{triades}-- built with a root note for each degree of a scale -- like
the ones in equation~\ref{eq:escalas}. In this context, music harmony aims at studying the observation of incident chord progressions and chaining rules. Even a monophonic melody generates harmonic fields, making it possible to observe the suggested chords at each passage.

In the 'traditional tonal harmony', a scale has its corresponding tonic (first scale
degree) and can be major (with the same notes of the Ionian mode) or
minor (notes from Eolian mode form the 'natural minor', which has both a
harmonic and a melodic version, see equation~\ref{eq:escalasMenores}). The
scale is base on triads, each one with its root and representing a degree into
the scale: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$. 
To build triads, the third and the fifth note above the root is considered together with the root (or fundamental). It is possible to write $\hat{1},\hat{3},\hat{5}$ as the first degree chord, built by the first degree of scale and the center for a tonal music. The chords
of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor
scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are secondary. After
that, other degrees are considered. The traditional harmony comprises conventions and stylistic techniques to create chord chains, made in each scale degree~\cite{Harmonia}.

The 'functional harmony' ascribes functions to these three central chords and
tries to understand their use by means of these functions. The chord built under
the first degree is named the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its (role) function consisting of preserving a center, usually referred as a ``ground'' for the music. The chord built under the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is call for the tonic (we say that the dominant chord asks for a conclusion and this conclusion is usually the tonic chord). Thus, the dominant chord guides the music to the tonic. The triad built under the forth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
subdominant, respectively) and its basically function is to distance the music from the tonic. The process aims at confirming the tonic using tonic-subdominant-tonic chains which are
expanded by using other chords in different ways.

The remaining triads are associated to these three most important chords. In the major scale, the associated relative (relative tonic \textit{Tr}, relative
subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built with a
third below, and the associated counter-relative (counter-relative
tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the
counter-relative dominant \textit{Da}) is the triad built with a third above. In the
minor scale the same happens, but the triad with a third below is called
counter-relative (tA, sA) and the triad with one third above is called relative (tR,
sR). The precise functions and musical effects of these chords are
controversial. Table~\ref{tab:harmonia} shows the relation between the triads
built in each degree of the major scale.

@@

The dominant counter-relative should form a minor chord. It explains the change
in the forth degree by a semitone above $\hat{7}\#$. The diminished chord
$\hat{7},\hat{2},\hat{4}$, is generally considered a 'dominant seventh chord
with no root'~\cite{Koellheuteur}.
In the minor mode, there is the change in $\hat{7}$ by an ascending semitone to
make possible the existence of an unique semitone separating the tonic, and
making also possible the dominant (that should be major and goes to the tonic). In
this way, the dominant is always major, for both major and minor scale and,
because of that, even as a minor tone, the relative dominant remains a third
below and in the counter-relative, a third above.

@@

Each one of these chords can be confirmed and developed by performing their
dominant or individual subdominant, which is the chord based in the third formed
by a fifth above or by a fifth below, respectively. These dominants and individual
subdominants, in the same way, have also subdominants and individual dominants that
can be used. Given a tonality, any chord can occur, no matter
how distant it is from the harmonic field and from the notes within the
scale. The unique on condition is that the occurrence presents a coherent trajectory of
dominants and subdominants to the original tonality.

Mediants, or 'chromatic mediants', are present in each chord in two ways: the upper chromatic
mediant, formed by the root at the third of original chord; and the lower
chromatic mediant, formed by the fifth at the third of original chord. Both chords also are formed by a third, but with a chromatic alteration regarding
the original chord. If two chromatic alterations exist, i.e.\ two notes altered
by one semitone each regarding the original chord, the mediant is called
'doubly chromatic mediant'. Again, there are two forms for each chord: the upper form, with
a third in the fifth of the original chord; and the lower form, with a third in the root
of the original triad. It is interesting to observe that a major chord has upper chromatic mediants and upper doubly chromatic mediants. A minor chord has lower chromatic mediants and lower doubly chromatic mediants. This relation between chords is considered as
advanced tonalism, sometimes even considered as expansion and dissolution of tonalism, and, as consequence, it has strong and outstanding effects although perfectly consonant. The chromatic
mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss,
among others, and are quite simple to be realized~\cite{Harmonia,Salzer}.

>>>>

The simultaneous occurrence of three or more notes is observed by means of chords.
Chords are based on triads, specially in tonal music. Triads are built by two successive thirds
within 3 notes: root, third and fifth. If the lower note of a chord is different from the root, this chord is 
an inverted chord as oposed to the chord in its root position. A closed position is any in which no chord note 
fits between two consecutive
notes~\cite{Lacerda}, any non-closed position is an open position. In closed and fundamental position,
and with fundamental in $0$, triads are described as:

@@

To have another third superimposed to the fifth, is sufficient to add $10$
as the highest note in order to form a tetrad with minor seventh, or add $11$ in order to form a tetrad with major
seventh. Inversions and open positions can be obtained with the 
addition of $\pm 12$ to selected component.

@@

For general guidance:

\begin{itemize}
        \item A fifth confirms the root (fundamental).
        \item Major or minor thirds, suggests major or minor chord qualities.
        \item Every tritone, specially if built between a major third and a
        minor seventh, tends to resolve in a third or a sixth.
        \item Note duplication is avoided. If duplication is needed, the
        preferred order is: the root, fifth, third and seventh.
        \item Note omission is avoided in the triad. If needed, the fifth is
        first considered for omission, than third and, if needed, the fundamental.
        \item It is possible to build chords with notes different from triads,
        specially if they respect a recurrent logic or musical concatenation that
        justifies these different notes.
        \item Chords built by successive intervals different of thirds -- like
        fourths and seconds -- are recurrent in compositions of advanced
        tonalism or experimental character.
        \item The repetition of chord successions (or their characteristics)
        fixes a trajectory and makes possible the
        introduction of exotic formations without incoherence.
\end{itemize}

@@

Omission of basic tonal chaining is the key to obtain modal and atonal harmonies. In
absence of these minimal tonal structures, the
harmony is considered modal if the notes match with some diatonic scale (see
equations~\ref{eq:escalas}) or if notes are presented in a small number.
If basic tonal chainings are absent and notes does not match any diatonic scale
and are diverse and dissonant (by relation with each other) enough to avoid
reduction by polarization, the harmony is atonal. In this classification, the
modal harmony is not tonal or atonal and is reduced to the 
incidence of notes within the diatonic scale (or simplifications) and to the
absence of tonal structures. It is possible to note, following this concept, that
atonal harmony is hard to be realized and, indeed, no matter how dissonant and diverse
is a set of notes, tonal harmonies arise if not avoided~\cite{harmEXT}.

@@

In fact, the techniques around atonal music aim at structures for avoiding
direct relation with modes and tonality. Manifesting such structures
are of such difficulty that the dodecafonism arouse. 
The purpose of dedocafonism is to use a set of notes (ideally 12 notes),
and to perform each note, one by one, in the same
order. In this context, the tonic becomes difficult to be stablished. Nevertheless, the
Western hearing searches for tonal traces in music and easily finds them by
unexpected and tortuous paths.

The use of dissonant intervals (specially tritones) without resolution,
reinforces the absence of tonality. Considering
this context, while creating a music
piece, it is possible to:

\begin{itemize}
     \item Repeat notes. By considering immediate repetition as an extension
     of previous incidence, the use of same note in sequence does not add relevant information.
     \item To play adjacent notes at the same time, making harmonic intervals
     and chords.
     \item Present durations and pauses with liberty, respecting notes order.
     \item Vary by extension, transposition, translation, inversion, retrograde and retrograde inversion. See subsections~\ref{subsec:motivos}
     and~\ref{subsec:usosmusicais3} for more details.
     \item Account for the presence of note structures, it is possible to use variations in orchestration, articulation, spatialization, among others.
\end{itemize}

The atonal harmony can be observed, paradigmatically, within these presented
conditions (a simple dodecaphonic model). Most of what was written by great dodecafonic composers, 
like Alban Berg and even Schoenberg, had the purpose of mixing tonal and atonal techniques.

@@

In the XX century, rhythmic music and with emphasis on sonorities/timbres, extended
the concepts of tonality and harmony. Hence, 
tonal harmony has strong influence in artistic movements and commercial
venues. In addition, the dodecafonism itself is considered of tonal nature because it
denies tonal characteristics of polarization.

In tonal or modal music, chords -- like the ones listed in
equations~\ref{triades}-- built on the root note of each degree of a scale  -- such 
as listed in equations~\ref{eq:escalas} --  form the pilars of the harmonic field. 
Music harmony aims at studying the incidence of chord progressions and chaining rules. 
Even a monophonic melody generates harmonic fields, making it possible to observe the suggested chords at each passage.

In the traditional tonal music, a scale has its tonic (first
degree) on any note, and can be major (with the same notes of Ionian mode) or
minor (same notes as Eolian mode, 'natural minor', which has both
harmonic and melodic versions, see equations~\ref{eq:escalasMenores}). The
scale is the base for triads, each with its root in a degree: $\hat{1},\hat{2},\hat{3},\hat{4},\hat{5},\hat{6},\hat{7}$. 
To build triads, the third and the fifth note above the root is considered together with the root (or fundamental).
$\hat{1},\hat{3},\hat{5}$ is the first degree chord,
 built on top of scale's first degree and central for tonal music. The chords
of the fifth degree $\hat{5},\hat{7},\hat{2}$ ($\hat{7}$ sharp when a minor
scale) and of the forth degree $\hat{4},\hat{6},\hat{1}$ are secondary. After
these, other degrees are considered. The `traditional harmony' comprises conventions
 and stylistic techniques to create chainings with such chords~\cite{Harmonia}.

The `functional harmony' ascribes functions to these three central chords and
tries to understand their use by means of these functions. The chord built on top of
first degree is the \textbf{tonic} chord (\textit{T} or \textit{t} for a major or minor tonic, respectively) and its function (role) consists of maintaining a center, usually referred as a ``ground'' for the music. The chord built on the fifth degree is the \textbf{dominant} (\textit{D}, the dominant is always major) and its function is to lean for the tonic
(the dominant chord asks for a conclusion and this conclusion is the tonic). Thus, the dominant chord guides the music to the tonic. The triad built under the forth degree is the \textbf{subdominant} (\textit{S} or \textit{s} for a major or minor
subdominant, respectively) and its function is to deviate the music from the tonic. 
The process aims at confirming the tonic using tonic-dominant-tonic chains which are
expanded by using other chords in different ways.

The remaining triads are associated to these three most important chords.
 In the major scale, the associated relative (relative tonic \textit{Tr}, relative
subdominant \textit{Sr} and relative dominant \textit{Dr}) is the triad built a
third below, and the associated counter-relative (counter-relative
tonic \textit{Ta}, counter-relative subdominant \textit{Sa} and the
counter-relative dominant \textit{Da}) is the triad built in a third above. 
In the minor scale the same happens, but the triad a third below is called
counter-relative (tA, sA) and the triad a third above is called relative (tR,
sR). The precise functions and musical effects of these chords are
controversial. Table~\ref{tab:harmonia} shows relations between the triads
built in each degree of the major scale.

@@

The dominant counter-relative should form a minor chord. It explains the change
in the forth degree by a semitone above $\hat{7}\#$. The diminished chord
$\hat{7},\hat{2},\hat{4}$, is generally considered a `dominant seventh chord
with omissed root'~\cite{Koellheuteur}.
In minor mode, there is a change in $\hat{7}$ by an ascending semitone to
make possible the existence of an unique semitone separating $\hat{7}$ and $\hat{1}$,
and making the dominant possible (that should be major and goes to the tonic). In
this way, the dominant is always major, for both major and minor scale and,
because of that, even in a minor tone, the relative dominant remains a third
below and in the counter-relative, a third above.

@@

Each one of these chords can be confirmed and developed by performing their
individual dominant or subdominant, which are the triads based
a fifth above or a fifth below, respectively. These individual dominants and subdominants,
in the same way, have also subdominants and individual dominants of their own.
Given a tonality, any chord can occur, no matter
how distant it is from the harmonic field and from the notes within the
scale. The unique condition is that the occurrence presents a coherent trajectory of
dominants and subdominants to the original tonality.

Mediants, or 'chromatic mediants', are two for each chord: the upper
mediant, formed by the root at the third of original chord; and the lower
mediant, formed by the fifth at the third of original chord.
Both chords also are formed by a third, but with a chromatic alteration regarding
the original chord. If two chromatic alterations exist, i.e.\ two notes altered
by one semitone each regarding the original chord, it is a
'doubly chromatic mediant'. Again, there are two forms for each chord: the upper form, with
a third in the fifth of the original triad; and the lower form, with a third in the root
of the original triad. It is interesting to observe that a major chord has major chromatic
 and doubly chromatic mediants. A minor chord has minor chromatic and doubly chromatic mediants.
(Recall that relatives and counter-relatives have oposite major-minor quality.)
This relation between chords is considered
advanced tonalism, sometimes even considered expansion and dissolution of tonalism,
with strong and impressing effects although perfectly consonant triads. Chromatic
mediants are used since the end of Romanticism by Wagner, Lizt, Richard Strauss,
among others, and are quite simple to be realized~\cite{Harmonia,Salzer}.
